The impacts of park access on health outcomes: a spatial comprehensive approach

### 2019-01-04 manuscript.Rnw
Rewrite results to focus on physical activity
Physical activity is what we actually hope will happen with parks. Let's look at obesity as a secondary outcome.

\maketitle
<<setup, cache=FALSE, echo=FALSE, message=FALSE>>=
library(tidyverse)
library(geojsonio)
library(rgdal)
library(rgeos)
library(maptools)
library(spdep)
library(sf)
library(VGAM)
library(qwraps2)
library(knitr)
library(kableExtra)
opts_chunk$set(cache=TRUE, message = FALSE, echo = FALSE)
library(leaflet)

@@

Consider that an individual is choosing a park for a recreation activity.
If we apply random utility choice theory, specifically the multinomial
logit model \citep{McFadden1974}, the expected probability of individual 
$i$ choosing park $j$ from the set of all regional parks $P$ is
\begin{equation}\label{eq:mnl}
P(j | V_{ij}) = \frac{\exp(V_{ij})}{\sum_{p \in P}\exp(V_{ip})}
\end{equation}
where parks are differentiated from each other by their relative measurable
utilities $V_{ij}$. In principle, $V$ may include any measurable attributes of either
the choice maker or the park. In this study we use a linear formulation of
\begin{equation}\label{eq:v_sd}
V_{ij} = size_j\beta_s + distance_{ij}\beta_d
\end{equation}
incorporating the size of the park in acres and the distance of the park from
the census tract $i$ in miles. The coefficients $\vec{\beta}$ can be 
estimated from household surveys, though in the absence of a survey we 
may assert reasonable values.

A key theoretical understanding of random utility choice models is that the
consumer surplus, or the total value of all alternatives of the choice set, can 
be obtained as the log-sum of the denominator of the choice probability equation:
\begin{equation}\label{eq:cs}
CS_i = \ln\left({\sum_{p \in J}\exp(V_{ip})}\right)
\end{equation}
There are several advantages to a log-sum defined metric relative to
buffer-based accessibility metrics more commonly found in the literature. First,
all individuals are defined as having some access to all parks, rather than an
arbitrary cutoff asserted by the researcher. This allows for the fact that some 
people are more or less sensitive to distances, and that distance is a 
continuous, and not a binary, phenomenon. Second, the random utility 
formulation allows the researcher to include any attribute of the park; 
in this case, we consider the
size of the park as an element of accessibility. This suggests that not all
parks are equal, and that a large park such as Central Park on Manhattan may
provide health and activity benefits over a much larger area than a smaller
community square.

Note that this formulation is an extension of the gravity-based accessibility
statistics, in the same way that the gravity model is a specific case of
a destination-choice model \citep{Daly1982}. The extension allows us to consider
more attributes of the park or the indivdual beyond size and impedance. For example,
we can aggregate the number of tweets emanating from a park as a measure of the
park's attractiveness beyond its size, and the utility for the choice maker from 
Equation \ref{eq:v_sd} then becomes
\begin{equation}\label{eq:v_sdt}
V_{ij} = size_j\beta_s + distance_{ij}\beta_d + tweets_j\beta_t
\end{equation}

\section{Empirical Application}

@@

<<parks>>=
parks <- sf::st_read("data/nyc_parks.geojson", quiet = TRUE) %>%
  transmute(
    size = Park_Acres,
    log_size = log(Park_Acres),
    tweets = TWEET_COUNT, 
    log_tweets = yeo.johnson(TWEET_COUNT, 0)
  )
@

We also collected data on park geometry.

\subsection{Model}

We predict the obesity rate in a census tract as a function of the tract's
sociodemographic characteristics and choice-based accessibility to parks with a
semi-log specification:
\begin{equation}\label{eq:themodel}
\log(\mathit{Obesity})_i = f(\vect{x}_i, CS_i, g(W, \mathit{Obesity}, \vect{X}))
\end{equation}
<<base_formula>>=
base_formula <- formula( 
  ~ log(density) + 
    fulltime + college + single + 
    Pct0to17 + Pct18to29 + Pct65plus + # need to leave out a category for collinearity  
    black + asian + hispanic + other)
@
where $\vect{x}_i$ is a vector of tract-level attributes, $CS_i$ is a choice-based
accessibility measure for the tract computed from Equations \ref{eq:v_sd} or 
\ref{eq:v_sdt}, and $g(W, \mathit{Obesity}, T)$
represents the impact of spatial neighborhood effects on the outcome.
Using tracts as the unit of observation in the model allows us to consider the
availability of parks at a metropolitan level, but it also introduces the problem
of spatial correlation of unobserved errors and spatial dependence in the data
generating process \citep{Cliff1970, Anselin1980}. Spatial correlation occurs
when spatially-distributed unobservable or missing variables (school quality,
neighborhood prestige, etc.) influence the modeled outcomes. Spatial dependence
occurs when the outcome {\em depends} on the outcome in neighboring areas; for
example, seeing neighbors exercising may encourage exercise. The relaxed spatial
Durbin model (SDM) proposed by \citet{Burridge1981} controls for both processes:
\begin{equation}\label{eq:sdm}
  \vect{y}= \rho W\vect{y} + X\vect{\beta} + WX\vect{\gamma} + \vect{\epsilon}
\end{equation}
where $\rho$ is the estimated spatial correlation between the outcome variables
$y_i, y_j, \ldots  \in \vect{y}$, $W$ is a matrix specifying the spatial
relationship between observations $i$ and $j$ in the dataset,
$\vect{\beta}$ is a vector of estimable coefficients relating the attributes $X$
of an observation to its outcome, and $\vect{\gamma}$ is a vector of
estimated coefficients relating the attributes of an observation's {\em neighbors}
to its outcome.

@@

\subsection{Accessibility}

@@

We initially estimated potential values of the $\beta$ coefficients by
applying a limited-memory, box-constrained optimization algorithm maximizing the
SDM model log-likehood. The constraint required that the coefficient on
size be positive and that on distance negative; all else equal, we assert people
will prefer to use larger parks and parks that are nearer to their residence. This
resulted in $\beta_d = -0.42$  and $\beta_s = 0.000$. Given our expectation that
park size is a valuable attribute, we subsequently adjusted this value to

@@

<<leaflet, eval=FALSE>>=
# this interactive map is much easier to create and explore
pal <- colorQuantile("Spectral", tracts_sf$access_ls, n = 5)
leaflet(tracts_sf) %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addPolygons(group = "access", color = ~ pal(access_ls), 
              label = ~as.character(round(access_ls,2))) %>% 
  addPolygons(group = "tweets", color = ~ pal(tweets_ls), 
              label = ~as.character(round(tweets_ls,2))) %>%
  addLayersControl(baseGroups = c("access", "tweets"))
@

@@

<<map, echo=FALSE, message=FALSE, warning=FALSE>>=
library(ggmap)
map <- ggmap(get_map(unname(st_bbox(tracts_sf)), source = "stamen", maptype = "toner"),
             extent = "device")
map +
  geom_sf(data = tracts_sf, inherit.aes=FALSE, 
          aes(fill = cut(access_ls, quantile(access_ls))), alpha = 0.8) +
  scale_fill_brewer("Relative Park Logsum", palette = "GnBu")
@
  \caption{Relative surplus-based accessibility in New York City}
  \label{fig:map}
\end{figure*}

@@

<<spatial selection>>=
obese_base_sem <- errorsarlm(update(base_formula, log_obesity ~ .), 
                             data = tracts@data, listw = W, zero.policy = TRUE)
obese_base_sdm <- lagsarlm(update(base_formula, log_obesity ~ .), 
                           data = tracts@data, listw = W, zero.policy = TRUE, 
                           type = "mixed")
test_sdmsem <- lmtest::lrtest(obese_base_sem, obese_base_sdm)
@

We estimated the A likelihood ratio test rejects that the SEM is equivalent, so we use the SDM
only going forward.

<<access_sdm>>=
# estimate models
access_sdm <- update(obese_base_sdm, .~ . + access_ls)
tweets_sdm <- update(obese_base_sdm, .~ . + tweets_ls)
@

\begin{table*}
  \caption{Estimated Coefficients on Obesity}
  \label{fig:obeseity_models}
<<modelchart, results="asis">>=
library(texreg)
obesity_models <- list(Base = obese_base_sdm, `Access` = access_sdm, 
                       `Access + Tweets` = tweets_sdm)
texreg(obesity_models, digits = 4, table = FALSE, single.row = TRUE, 
       booktabs = TRUE, use.packages = FALSE)
@
\end{table*}

\section{Conclusion}

>>>>

\maketitle
<<setup, cache=FALSE, echo=FALSE, message=FALSE>>=
library(tidyverse)
library(geojsonio)
library(rgdal)
library(rgeos)
library(maptools)
library(spdep)
library(sf)
library(VGAM)
library(qwraps2)
library(knitr)
library(kableExtra)
library(texreg)
opts_chunk$set(cache=TRUE, message = FALSE, echo = FALSE)
library(leaflet)

@@

Consider that an individual is choosing a park for a recreation activity.
If we apply random utility choice theory, specifically the multinomial
logit model \citep{McFadden1974}, the expected probability of individual 
$i$ choosing park $j$ from the set of all regional parks $J$ is
\begin{equation}\label{eq:mnl}
P(j | V_{ij}) = \frac{\exp(V_{ij})}{\sum_{p \in J}\exp(V_{ip})}
\end{equation}
where parks are differentiated from each other by their relative measurable
utilities $V_{ij}$. In principle, $V$ may include any measurable attributes of either
the choice maker or the park. In this study we use a linear formulation of
\begin{equation}\label{eq:v_sd}
V_{ij} = size_j\lambda_s + distance_{ij}\lambda_d
\end{equation}
incorporating the size of the park in acres and the distance of the park from
the census tract $i$ in miles. The coefficients $\vect{\lambda}$ can be 
estimated from household surveys, though in the absence of a survey we 
may assert reasonable values.

A key theoretical outcome of random utility choice models is that the
consumer surplus, or the total value of all alternatives of the choice set, can 
be obtained as the log-sum of the denominator of the choice probability equation:
\begin{equation}\label{eq:cs}
CS_i = \ln\left({\sum_{p \in J}\exp(V_{ip})}\right)
\end{equation}
There are several advantages to a log-sum defined metric relative to
buffer-based accessibility metrics more commonly found in the literature. First,
all individuals are defined as having some access to all parks, rather than an
arbitrary cutoff asserted by the researcher. This allows for the fact that some 
people are more or less sensitive to distances, and that distance is a 
continuous, and not a binary, phenomenon. Second, the random utility 
formulation allows the researcher to include any attribute of the park; 
in this case, we consider the
size of the park as an element of accessibility. This suggests that not all
parks are equal, and that a large park such as Central Park on Manhattan may
provide health and activity benefits over a much larger area than a smaller
community square.

Note that this formulation is an extension of the gravity-based accessibility
statistics, in the same way that the gravity model is a specific case of
a destination-choice model \citep{Daly1982}. The extension allows us to consider
more attributes of the park or the indivdual beyond size and impedance. For example,
we can aggregate the number of tweets emanating from a park as a measure of the
park's attractiveness beyond its size, and the utility for the choice maker from 
Equation \ref{eq:v_sd} then becomes
\begin{equation}\label{eq:v_sdt}
V_{ij} = size_j\lambda_s + distance_{ij}\lambda_d + tweets_j\lambda_t
\end{equation}

\section{Empirical Methodology}

@@

<<parks>>=
parks <- sf::st_read("data/nyc_parks.geojson", quiet = TRUE) %>%
  transmute(
    size = Park_Acres,
    log_size = log(Park_Acres),
    tweets = TWEET_COUNT, 
    log_tweets = yeo.johnson(TWEET_COUNT, 0)
  ) %>% 
  filter(size > 0.5)
@

We also collected data on park geometry, limited to parks larger than 0.5 acres.

\begin{table*}
  \caption{Summary Statistics of Parks Data}
  \label{tab:parksdata}
<<parks_summary, results="asis", message=FALSE, warning=FALSE, eval = FALSE>>=
park_summary <- list(
  "size"     = tab_summary(parks$size),
  "tweets"   = tab_summary(parks$tweets)
)

summary_table(as_data_frame(parks), park_summary) %>%
  kable("latex", booktabs = T, escape = FALSE, col.names = c("")) %>%
  group_rows("Size: size of park in acres", 1, 4) %>%
  group_rows("Tweets: number of tweets emanating from park", 5, 8)
@
\end{table*}

\subsection{Model}\label{subsec:model}

We predict the rate of physical activity ($y$) in a census tract as a 
function of the tract's sociodemographic characteristics and choice-based 
accessibility to parks with a semi-log specification:
\begin{equation}\label{eq:themodel}
\log(y_i) = f(\vect{x}_i, CS_i, g(W, \mathit{Obesity}, \vect{X}))
\end{equation}
<<base_formula>>=
base_formula <- formula( 
  ~ log(density) + 
    fulltime + college + single + 
    Pct0to17 + Pct18to29 + Pct65plus + # need to leave out a category for collinearity  
    black + asian + hispanic + other)
@
where $\vect{x}_i$ is a vector of tract-level sociodemographic attributes, 
$CS_i$ is a choice-based
accessibility measure for the tract computed from Equation \ref{eq:cs} and 
$g(W, \vect{y}, X)$
represents the impact of spatial neighborhood effects on the outcome.

Using tracts as the unit of observation in the model allows us to consider the
availability of parks at a sub-metropolitan level, but it also introduces the problem
of spatial correlation of unobserved errors and spatial dependence in the data
generating process \citep{Cliff1970, Anselin1980}. Spatial correlation occurs
when spatially-distributed unobservable or missing variables (school quality,
neighborhood prestige, etc.) influence the modeled outcomes. Spatial dependence
occurs when the outcome {\em depends} on the outcome in neighboring areas; for
example, seeing neighbors exercising may encourage exercise. The relaxed spatial
Durbin model (SDM) proposed by \citet{Burridge1981} controls for both processes:
\begin{equation}\label{eq:sdm}
  \vect{y}= \rho W\vect{y} + X\vect{\beta} + WX\vect{\gamma} + \vect{\epsilon}
\end{equation}
where $\rho$ is the estimated spatial correlation between the outcome variables
$y_i, y_j, \ldots  \in \vect{y}$, $W$ is a matrix specifying the spatial
relationship between observations $i$ and $j$ in the dataset,
$\vect{\beta}$ is a vector of estimable coefficients relating the attributes $X$
of an observation to its outcome, and $\vect{\gamma}$ is a vector of
estimated coefficients relating the attributes of an observation's {\em neighbors}
to its outcome.

@@

\subsection{Accessibility}\label{subsec:accessibility}

@@

We initially estimated potential values of the $\beta$ coefficients by
applying a limited-memory, box-constrained optimization algorithm maximizing the
SDM model log-likehood. The constraint required that the coefficient on
size be positive and that on distance negative; all else equal, we assert people
will prefer to use larger parks and parks that are nearer to their residence. This
resulted in $\beta_d = -0.42$  and $\beta_s = 0.000$. Given our expectation that
park size is a valuable attribute, we subsequently adjusted this value to
a value of $1.00$. This implies that each mile

@@

<<leaflet, eval=FALSE>>=
# this interactive map is much easier to create and explore
pal <- colorQuantile("GnBu", tracts_sf$access_ls, n = 5)
leaflet(tracts_sf) %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addPolygons(group = "access", color = ~ pal(access_ls), 
              label = ~as.character(round(access_ls,2))) %>% 
  addPolygons(group = "tweets", color = ~ pal(tweets_ls), 
              label = ~as.character(round(tweets_ls,2))) %>%
  addLayersControl(baseGroups = c("access", "tweets"))
@

@@

<<map, echo=FALSE, message=FALSE, warning=FALSE>>=
library(ggmap)
map <- ggmap(get_map(unname(st_bbox(tracts_sf)), source = "stamen", maptype = "toner-lite"),
             extent = "device")
map +
  geom_sf(data = tracts_sf, inherit.aes=FALSE, 
          aes(fill = cut(access_ls, breaks = c(-Inf, -3, -2, -1, 0, 2, 3, Inf))),
          alpha = 0.8) +
  scale_fill_brewer("Relative Park Logsum", palette = "GnBu")
@
  \caption{Relative surplus-based accessibility in New York City}
  \label{fig:map}
\end{figure*}

@@

<<spatial selection>>=
physact_base_sem <- errorsarlm(update(base_formula, log_physact ~ .), 
                             data = tracts@data, listw = W, zero.policy = TRUE)
physact_base_sdm <- lagsarlm(update(base_formula, log_physact ~ .), 
                           data = tracts@data, listw = W, zero.policy = TRUE, 
                           type = "mixed")
test_sdmsem <- lmtest::lrtest(physact_base_sem, physact_base_sdm)
@

<<access_sdm>>=
# estimate models
pa_access_sdm <- update(physact_base_sdm, .~ . + access_ls)
pa_tweets_sdm <- update(physact_base_sdm, .~ . + tweets_ls)
@

We estimated SDM and SEM models regressing the physical activity rate against
against the base covariates (without the accessibility logsum). A likelihood 
ratio test rejects that the SEM is equivalent 
($\chi^2 = \Sexpr{test_sdmsem$Chisq[2]}$ on \Sexpr{test_sdmsem$Df[2]} degrees of 
freedom), so we use the SDM specification only going forward. Table 
\ref{tab:pa_models} presents the estimated coefficients for the base model as 
well as models including logsum-based accessibility with and without Twitter
activity. A few initial observations can be made from this table. First, adding
the logsum measurement significantly improves the overall model fit, using a
likelihood ratio test on the model log-likelihood statistics. Second, the
coefficient estimates on the controlling variables remain virtually unchanged 
from the base model.

\begin{table*}
  \caption{Estimated Coefficients Relating Park Access to Physical Activity}
  \label{tab:pa_models}
<<pamodelchart, results="asis">>=
pa_models <- list(Base = physact_base_sdm, `Access` = pa_access_sdm, 
                       `Access + Tweets` = pa_tweets_sdm)
texreg(pa_models, digits = 4, table = FALSE, single.row = TRUE, 
       booktabs = TRUE, use.packages = FALSE)
#' screenreg(pa_models, digits = 4, table = FALSE) 
@
\end{table*}

Recall from the presentation of spatial econometrics above that the total effect of a
covariate on the dependent variable is different from the estimated coefficient
and involves interaction between direct and indirect effects, or between 
the $\beta$ and $\gamma$ parameters. This is particularly important when the
$\beta$ and $\gamma$ estimates for a particular covariate show a different
sign, such as in the proportion of black residents in a census tract, as 
interpreting the direction and magnitude of the relationship is not straightforward.

<<extract_impacts>>=
#' Get the MC simulated impact coefficients and significance from a model
#' @param sdm An autoregressive lag (SAR or SDM) model object `lagsarlm`
#' @return A data frame with the effect (direct, indirect, total), variable, 
#'   simulated mean impact, and significance p-value
extract_impacts <- function(sdm) {

  impacts_summary <- summary(
    impacts(sdm, tr=trMC,  R = 1000, useHESS = !is.logical(sdm$fdHess)), 
    zstats = TRUE)

  estimate <- list(
    Direct = impacts_summary$direct_sum,
    Indirect = impacts_summary$indirect_sum,
    Total = impacts_summary$total_sum
  ) %>%
  lapply(function(s) {
    data_frame(
      term = names(s$statistics[,1]),
      estimate = s$statistics[,1],
      std.error = s$statistics[,2]
    )
  }) %>%
    bind_rows(.id = "effect")

  pval <- as_data_frame(impacts_summary$pzmat) %>%
    mutate(
     term = rownames(impacts_summary$pzmat)
    ) %>%
    gather("effect", "p-val", -term)

  left_join(estimate, pval, by = c("effect", "term"))
}
@

\begin{table*}
  \caption{Simulated Impact of Accessibility on Physical Activity}
  \label{tab:pa_impacts}
<<impacts>>=
physact_impacts <- pa_models %>%
  lapply(extract_impacts) %>%
  bind_rows(.id = "model") %>% ungroup() %>%
  rename(submodel = effect)

effect_estimate <- physact_impacts %>% 
  filter(model == "Access + Tweets", term == "tweets_ls", submodel == "Total") %>% 
  pull(estimate)

density_estimate <- physact_impacts %>% 
  filter(model == "Access + Tweets", term == "log(density)", submodel == "Total") %>% 
  pull(estimate)

physact_impacts %>%
  transmute(
    model, term, submodel, estimate,
    output = paste(round(`p-val`, 5), 
                   gtools::stars.pval(`p-val`))
  ) %>%
  filter(term %in% c("access_ls", "tweets_ls")) %>%
  select(-term) %>%
  kable("latex", booktabs = T, escape = FALSE, 
        col.names = c("Logsum Measure", "Effect", "Estimate", "$p$-value"))
@
\end{table*}

Table \ref{tab:pa_impacts} shows the simulated marginal impact of the
accessibility logsum measurements on the physical activity rate in each tract.
Given that the logsum is a relatively continuous measure over space, and that it
would be difficult to improve the park accessibility of a tract without also 
improving the accessibility of the tract's neighbors, the total effect combining
the direct effect and the indirect effects of all neighbors is 
most appropriate. The total effect of tract-level accessiblity on physical 
activity rates is strongly significant, regardless of whether we consider Twitter
activity.

It is of course necessary to also consider the magnitude of the effect. 
As this is a semi-log model, the coefficient implies a unit increase
in the access and tweets logsum leads to a 
\Sexpr{round(effect_estimate * 100, 3)}\% increase in the physical activity rate. 
Referring to the map of logsum measures in Figure \ref{fig:map}, the measures
span from roughly $-3$ in heavily industrialized segments of southern Brooklyn
to roughly $3$ in tracts immediately adjacent to Central Park on the Upper East
Side. Thus moving from the least accessible parts of the city to the most
accessible will reduce the expected physical activity rate by 
\Sexpr{round(effect_estimate * 100 * 6, 3)}\%. For an aggregate statistic, this 
is not an insubstantial marginal effect.

\section{Discussion}
